---
title: "Data621_Project"
author: "Saloua Daouki"
date: "2024-12-10"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## # 2023-2024 Science Benchmark Data:

**Research Question:** What are the trends in student performance on science benchmark assessments among 6th, 7th, and 8th graders at a Brooklyn school over the academic year? How can regression modeling help identify and forecast performance patterns, highlighting specific questions or standards that show consistent struggles or growth across three benchmarks, and how accurately can these trends predict future performance?

## 1. Load necessary libraries:

This part sets the foundation for the analysis, making sure all necessary tools are in place before moving on to data cleaning, exploration, and visualization. Without loading these libraries, the subsequent analysis wouldn’t be possible, so it's a crucial preparatory step.

```{r }
# Load necessary libraries for data analysis and visualization
library(readxl)   # For reading Excel files
library(purrr)
library(dplyr)    # For data manipulation and transformation
library(tidyr)    # For tidying data
library(naniar)   # For visualizing and handling missing values
library(visdat)   # For creating heatmaps for missing values
library(ggplot2)  # For data visualization
library(DT)       # For creating interactive data tables in R
library(reshape2) 
library(knitr)
library(corrplot)  # For correlation plots
library(stringr)  # Load stringr for string extraction
library(gt)
library(plotly) #allow the plots to be zoomed into for better inspection without sacrificing plot quality.
```

## 2. Explore the data

### 2.1 Load the data

This section establishes a solid base for my future exploration, cleaning, and analysis of student performance trends. I am now prepared to dive deeper into identifying trends and insights across benchmarks.

```{r}
# Path to the Excel file
file_path1 <- "2023 - 2024 SCIENCE BENCHMARK DATA  copy.xlsx"
```

- **Reading 6th, 7th, and 8th Grade Benchmarks:**

  - `The read_excel()` function is used to load the data for different grade levels and benchmarks. This includes data from three assessments for each grade, allowing us to analyze performance trends over time.

  - The benchmark sheets are labeled accordingly (e.g., "6th Grade Benchmark 1", "7th Grade Benchmark 2"), and the data is read for each grade level separately.
  
- **Cleaning 7th Grade Data:**

Here, I am importing the benchmark data for 6th, 7th, and 8th grades from an Excel file, pulling in data for each of the three benchmarks administered over the academic year.

```{r}
# Load data for each grade and benchmark
# Reading 6th grade benchmark data
benchmark_6_11 <- read_excel(file_path1, sheet = "6th Grade Benchmark 1")
benchmark_6_22 <- read_excel(file_path1, sheet = "6th Grade Benchmark 2")
benchmark_6_33 <- read_excel(file_path1, sheet = " 6th Grade Benchmark 3")

# Reading 7th grade benchmark data, while removing specific unnecessary (hidden) columns
benchmark_7_11 <- read_excel(file_path1, sheet = "7th Grade Benchmark 1")
benchmark_7_22 <- read_excel(file_path1, sheet = "7th Grade Benchmark 2")
benchmark_7_33 <- read_excel(file_path1, sheet = "7th Grade Benchmark 3")

# Reading 8th grade benchmark data
benchmark_8_11 <- read_excel(file_path1, sheet = "8th Grade Benchmark 1")
benchmark_8_22 <- read_excel(file_path1, sheet = "8th Grade Benchmark 2")
benchmark_8_33 <- read_excel(file_path1, sheet = "8th Grade Benchmark 3")
```

### 2.2 Load all data into a list:

This section is about optimizing the way we handle multiple datasets by loading them into a list structure. This approach provides several benefits, especially for simplifying future data manipulation and ensuring consistency.

```{r}
# Load data from all sheets into a list
sheet_names1 <- c("6th Grade Benchmark 1", "6th Grade Benchmark 2", " 6th Grade Benchmark 3",
                 "7th Grade Benchmark 1", "7th Grade Benchmark 2", "7th Grade Benchmark 3",
                 "8th Grade Benchmark 1", "8th Grade Benchmark 2", "8th Grade Benchmark 3")

# Load data with map to streamline the process
benchmarks_list1 <- map(sheet_names1, ~ read_excel(file_path1, sheet = .x))
names(benchmarks_list1) <- sheet_names1
```

The repeated "New names" in the message above indicates an issue with column naming, similar to the previous section. It might be related to renaming variables or cleaning the column names, which will be done later once all the data is loaded into the list.

### 2.3 Examine the data structure 

The goal here is to quickly inspect the first few rows of each dataset in benchmarks_list to confirm that the data has been loaded correctly and to get an initial sense of the structure.

```{r}
# Inspect the first few rows of the benchmarks list to ensure data is loaded
lapply(benchmarks_list1, head)
```


The list structure ($-sign notation) shows that each dataset is being listed by its sheet name (e.g., $6th Grade Benchmark 1, $7th Grade Benchmark 2`, etc.).

The other outputs are the first few rows of each dataset, displaying the varaibles for each.

```{r}
# Examine the structure of the first benchmark data
str(benchmarks_list1[[9]])  # Display the structure of the ninth sheet
```

It appears the `% Correct` row is part of the data and will interfere with further analysis. I'll address this issue in the next section by cleaning each tab to prepare for proper visualization and exploration.

- **Key Observations on Data Structure:**

    - Column Headers: The first column, which should represent question numbers, lacks a clear label. Renaming it to "Questions" will improve clarity.   
    
    - Subheadings: The remaining columns, currently labeled "% Correct," need to be renamed to reflect the class numbers (e.g., "600," "817") for accurate analysis and easy reference.
    
    - Tidy Data: Ensuring each variable has its own column and each observation its own row, following tidy data principles, will streamline data manipulation and reduce errors during analysis.

### 2.4 Visualize Missing Values:

In this section, we assess data quality by visualizing missing values across the benchmark datasets. Identifying missing data early on helps determine if any imputation or cleaning is necessary to ensure reliable analysis.

```{r}
# Visualize missing values in each benchmark dataset with titles
vis_miss(benchmark_6_11) + ggtitle("6th Grade Benchmark 1")
vis_miss(benchmark_6_22) + ggtitle("6th Grade Benchmark 2")
vis_miss(benchmark_6_33) + ggtitle("6th Grade Benchmark 3")

vis_miss(benchmark_7_11) + ggtitle("7th Grade Benchmark 1")
vis_miss(benchmark_7_22) + ggtitle("7th Grade Benchmark 2")
vis_miss(benchmark_7_33) + ggtitle("7th Grade Benchmark 3")

vis_miss(benchmark_8_11) + ggtitle("8th Grade Benchmark 1")
vis_miss(benchmark_8_22) + ggtitle("8th Grade Benchmark 2")
vis_miss(benchmark_8_33) + ggtitle("8th Grade Benchmark 3")
```

**Findings:**

  - 7th Grade Benchmark 2: One column has 96% missing values, indicating potential issues with data collection or formatting.
  
  - 8th Grade Benchmark 3: Twelve columns contain 96% missing values, with 43.5% of the dataset being present and 56.5% missing. This could significantly affect the analysis if not addressed.
  
  - Other Datasets: No missing values are detected in the remaining datasets.

### 2.5 Explore Each Tab: 

In this section, we iterate through the first three benchmark datasets (6th Grade Benchmark 1, 6th Grade Benchmark 2, 6th Grade Benchmark 3) and perform the following actions:

- Remove Redundant Rows: The first row, which contains '% Correct,' is removed as it does not represent relevant data.

- Rename Columns: The first column is renamed to "Questions," while the last two columns are renamed "Per_Question_Average" and "Skills" for clarity and consistency.

```{r}
# Loop through the first three data frames in benchmarks_list to remove the first row and rename columns
for (i in 1:3) {  # Only iterate through the first three tabs
  df1 <- benchmarks_list1[[i]]
  
  # Remove the first row containing "% Correct"
  df1 <- df1[-1, ]
  
  # Rename the first column
  colnames(df1)[1] <- "Questions"  
  
  # Rename the second-to-last and last columns
  last_col1 <- ncol(df1)
  colnames(df1)[last_col1 - 1] <- "Per_Question_Average"  # Second-to-last column
  colnames(df1)[last_col1] <- "Skills"  # Last column
  
  # Update the list with the modified data frame
  benchmarks_list1[[i]] <- df1
}

# Check the updated structure of the data frames
str(benchmarks_list1[[1]])
```

We did the same process as previous in the second set of benchmark datasets (4th to 6th in the list, corresponding to the 7th Grade Benchmark 1, 7th Grade Benchmark 2, and 7th Grade Benchmark 3).

```{r}
# Loop through the specified data frames in benchmarks_list to remove the first row if necessary and rename columns
for (j in 4:6) {  # Only iterate through the second three tabs
  df1 <- benchmarks_list1[[j]]
  
  # Check if the first row contains "% Correct" and only then remove it
  if (any(grepl("% Correct", df1[1, ], fixed = TRUE))) {
    df1 <- df1[-1, ]
  }
  
  # Rename columns
  colnames(df1)[1] <- "Questions"  # Rename the first column
  
  # Rename the last column as "Per_Question_Average"
  last_col1 <- ncol(df1)
  colnames(df1)[last_col1] <- "Per_Question_Average"
  
  # Update the list with the modified data frame
  benchmarks_list1[[j]] <- df1
}

# Optional: Check the updated structure of the data frames
str(benchmarks_list1[[4]])
```

Again, here we are cleaning and renaming the last three benchmark datasets (8th Grade Benchmark 1, 8th Grade Benchmark 2, and 8th Grade Benchmark 3)

```{r}
# Loop through the first three data frames in benchmarks_list to remove the first row and rename columns
for (k in 7:9) {  # Only iterate through the last three tabs
  df1 <- benchmarks_list1[[k]]
  
  # Remove the first row containing "% Correct"
  df1 <- df1[-1, ]
  
  # Rename columns
  colnames(df1)[1] <- "Questions"           # Rename the first column
  last_col1 <- ncol(df1)
  colnames(df1)[last_col1] <- "Per_Question_Average"
  
  # Update the list with the modified data frame
  benchmarks_list1[[k]] <- df1
}

# Optional: Check the updated structure of the data frames
str(benchmarks_list1[[7]])
```

By applying these changes, the datasets now have a cleaner, more uniform structure, making further analysis easier and more accurate.

Next, we implement a crucial step to ensure the integrity of our data before proceeding with further analysis: checking for any empty data frames within our benchmarks_list.

The primary goal is to identify any data frames that may not contain any rows (i.e., are empty). This is an important quality control step, as empty data frames could indicate issues with data loading or data extraction processes. These empty frames would not provide any useful information for our analysis and could lead to errors if not addressed.

```{r}
# Check if any data frames are empty
empty_checks1 <- sapply(benchmarks_list1, function(df1) nrow(df1) == 0)
print(empty_checks1)  # This will show TRUE for any empty data frame
```
The printed output above shows that none of the benchmark datasets are empty. it indicates that all datasets contain data and are ready for further analysis.

The next exploration section plays a crucial role in both understanding the datasets and preparing for more detailed analyses. It highlights not only the overall performance trends but also areas where students may need additional support.

```{r}
# Define a function to explore the dataset
explore_data1 <- function(df1) {
  cat("### First Few Rows of the Data ###\n")
  print(head(df1))
  
  cat("\n### Structure of the Data ###\n")
  print(str(df1))
  
  cat("\n### Summary Statistics of the Data ###\n")
  print(summary(df1))
  
  # Check if the column 'Per_Question_Average' exists
  if ("Per_Question_Average" %in% colnames(df1)) {
    # Ensure the column is numeric for plotting if necessary
    df1$Per_Question_Average <- as.numeric(df1$Per_Question_Average)

    # Visualization: Histogram for Per_Question_Average
    if (!all(is.na(df1$Per_Question_Average))) { # Check if the column has non-NA values
      cat("\n### Histogram of Per_Question_Average ###\n")
      print(
        ggplot(df1, aes(x = Per_Question_Average)) +
          geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
          labs(title = "Distribution of Per_Question_Average", x = "Per Question Average", y = "Frequency") +
          theme_minimal()
      )
    } else {
      cat("\nPer_Question_Average has only NA values.\n")
    }

    # Visualization: Bar Plot for Per_Question_Average by Questions
    if (!all(is.na(df1$Per_Question_Average))) { # Check if the column has non-NA values
      cat("\n### Bar Plot of Per_Question_Average by Question ###\n")
      
      # Ensure that 'Questions' column is ordered as factors from 1 to 25
      df1$Questions <- factor(df1$Questions, levels = 1:25)

      print(
        ggplot(df1, aes(x = Questions, y = Per_Question_Average)) +
          geom_bar(stat = "identity", fill = "lightgreen", color = "black", na.rm = TRUE) +
          labs(title = "Per Question Average by Question", x = "Questions", y = "Per Question Average") +
          theme_minimal() +
          theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for readability
      )
    } else {
      cat("\nPer_Question_Average column has only NA values.\n")
    }
  } else {
    cat("\nThe column 'Per_Question_Average' does not exist in this dataset.\n")
  }
}

# Apply the function to each benchmark
for (i in 1:length(benchmarks_list1)) {
  cat("\n\n### Exploring Benchmark", i, "###\n")
  explore_data1(benchmarks_list1[[i]])
}
```

- The histogram (blue graphs) helps to see the overall spread of average scores across all questions by showing how frequently each score range appears. This the question that is answered by the histogram: **How often do different average scores appear?**

- The bar plot (green plots) provides a direct comparison of individual question scores, helping identify which questions scored higher or lower on average. This is the question that is answered by the bar plot: **What is the average score for each question?**

## 3. Data Cleaning and Preparation

### 3.1 Preparing the data:

In this step, we inspected the dataset to confirm the presence of the necessary columns. The dataset (8th grade as an example) contains the following columns:

```{r}
# Check if the necessary columns are present before selection
print(paste("Columns before selection:", paste(colnames(df1), collapse=", ")))
```

By reviewing this structure, we ensure that all the necessary columns are available for further analysis, which includes comparing performance across different benchmarks and identifying trends in student performance for each question using the `Per_Question_Average`.

Next, a check has been performed for non-numeric entries in the benchmark data frames, and we've found some entries that are not suitable for the analysis.

```{r}
# Function to check for non-numeric entries
check_non_numeric1 <- function(df1) {
  sapply(df1, function(col) {
    if (is.character(col)) {
      # Check for unique non-numeric values
      unique_values1 <- unique(col[!is.na(col) & col != ""])
      non_numeric_values1 <- unique_values1[!grepl("^[0-9.]+$", unique_values1)]
      return(non_numeric_values1)
    } else {
      return(NULL)
    }
  })
}

# Apply the function to each benchmark data frame
non_numeric_check1 <- lapply(benchmarks_list1, check_non_numeric1)
print(non_numeric_check1)
```

**Findings**

- Common Issues Across All Benchmarks:

    - Questions Column: Contains "Class Average" (for each benchmark).

    - Per_Question_Average Column: Contains "Grade Average" (for each benchmark).

    - Specific Class Columns: Some columns (like 621 in 6th Grade Benchmark 1) contain entries like "Class has 6 students."

- Skills Column: Contains non-numeric values that describe the skills associated with the questions.

To clean up these non-numeric entries, I am going to Convert Per_Question_Average to Numeric in Each Data Frame. This will ensure that the Per_Question_Average column in each benchmark data frame is numeric before combining them.

```{r}
# Ensure Per_Question_Average is numeric for all data frames
benchmarks_list1 <- lapply(benchmarks_list1, function(df1) {
  if ("Per_Question_Average" %in% colnames(df1)) {
    # Convert non-numeric values to NA, ignoring warnings
    df1$Per_Question_Average <- suppressWarnings(as.numeric(as.character(df1$Per_Question_Average)))
  }
  return(df1)
})

# Print a sample to check if conversion worked correctly
print(benchmarks_list1[[1]]$Per_Question_Average)
```

### 3.2 Cleaning the Data:

This step aims to clean up the dataset by handling missing values, converting columns to numeric types, and ensuring consistency across different benchmark tabs.

```{r}
clean_tab1 <- function(data1, sheet_name1) {
  # Debug: Print the data frame before cleaning
  cat("Before cleaning:", sheet_name1, "\n")
  print(head(data1))

  # Replace "???" and empty strings with NA for all character columns
  data1 <- data1 %>%
    mutate(across(where(is.character), ~ na_if(na_if(., "???"), "")))

  # Debug: Print the data frame after replacing "???"
  cat("After replacing '???':", sheet_name1, "\n")
  print(head(data1))

  # Convert character columns to numeric where applicable, excluding the 'Skills' column
  data1 <- data1 %>%
    mutate(across(where(is.character) & !any_of("Skills"), ~ suppressWarnings(as.numeric(.))))

  # Debug: Print the data frame after conversion
  cat("After conversion:", sheet_name1, "\n")
  print(head(data1))

  # Round numeric columns to 2 decimal places, excluding the 'Skills' column
  data1 <- data1 %>%
    mutate(across(where(is.numeric) & !any_of("Skills"), ~ round(., 2)))

  return(data1)
}

# Apply the cleaning function to each tab in benchmarks_list
benchmarks_list1 <- setNames(
  lapply(names(benchmarks_list1), function(sheet_name1) {
    clean_tab1(benchmarks_list1[[sheet_name1]], sheet_name1)
  }), names(benchmarks_list1))

# Display the first few rows of the cleaned 7th Grade Benchmark 1 data
head(benchmarks_list1[["7th Grade Benchmark 1"]])
```

**Before Cleaning:** Each debug message labeled "Before cleaning" shows the initial state of the data in each benchmark tab. It allows us to see if there are any anomalies, such as the presence of ??? or inconsistent formats before cleaning.

**After Replacing '???':** Each "After replacing '???'" message shows that any placeholder ??? values have been successfully replaced by NA. This ensures missing data is handled correctly.

**After Conversion:** Each "After conversion" message confirms that character columns have been converted to numeric where applicable, ensuring that you can perform calculations and analysis without errors due to character data types.

### 3.3 Checking the cleaned data:

When combining the data frames, check that the bind_rows function is receiving properly formatted data frames. If the data frames were cleaned to the point where they only contain NA or empty rows, the combined data will also reflect that.

```{r}
# Before combining, check if any cleaned data frames are empty
cleaned_checks1 <- lapply(benchmarks_list1, function(df1) nrow(df1) == 0)
print(cleaned_checks1)  # Should show FALSE for all
```

**No Empty Data Frames:**

Since every result is FALSE, it means that none of the benchmarks lost all their data during the cleaning process. This is a positive outcome, confirming that the cleaning steps—such as replacing ???, handling missing values, and converting columns to numeric—did not remove so much data that any benchmark tab became completely empty.

### 3.4 Adding categories to the dataset

```{r}
# Define class types for each grade level
class_types_6th <- tibble(
  Class = as.character(c(600:604, 605:608, 609:610, 611:620, 621, 673, 676)),
  Class_Type = c(
    rep("IIM", 5),             # Classes 600 to 604
    rep("ICT", 4),             # Classes 605 to 608
    rep("Self_contained", 2),  # Classes 609 & 610
    rep("General", 10),        # Classes 611 to 620
    "Dual Language",           # Class 621
    "Self_Contained",          # Class 673
    "Newcomers"                # Class 676
  )
)

class_types_7th <- tibble(
  Class = as.character(c(700:703, 704, 705:707, 708, 709:710, 711:714, 715, 716, 717:721)),
  Class_Type = c(
    rep("Regents", 4),         # Classes 700 to 703
    "Regents_ICT",             # Class 704
    rep("ICT", 3),             # Classes 705 to 707
    "General",                 # Class 708
    rep("Self_contained", 2),  # Classes 709 & 710
    rep("General", 4),         # Classes 711 to 714
    "Virtual_Enterprise",      # Class 715
    "IIM_Regents",             # Class 716
    rep("General", 5)          # Classes 717 to 721
  )
)

class_types_8th <- tibble(
  Class = as.character(c(800:803, 804, 805:808, 809:810, 811:814, 815, 816, 817:818)),
  Class_Type = c(
    rep("Regents", 4),           # Classes 800 to 803
    "General",                   # Class 804
    rep("ICT", 4),               # Classes 805 to 808
    rep("Self_contained", 2),    # Classes 809 to 810
    rep("General", 4),           # Classes 811 to 814
    "Virtual_Enterprise",        # Class 815
    "IIM",                       # Class 816
    rep("Other",2)               # Classes 817 & 818
  )
)

# Function to pivot long and assign class types based on grade
process_benchmark <- function(data, class_types) {
  data %>%
    pivot_longer(
      cols = -c(Questions, Per_Question_Average, Skills),   # Only pivot the class columns (leave the rest)
      names_to = "Class",                                     # Create a column for the class
      values_to = "Score"                                     # Values go into the "Score" column
    ) %>%
    left_join(class_types, by = "Class")                      # Add Class_Type based on the Class
}

# Apply the function to each grade's benchmarks in the list
benchmarks_list1[[1]] <- process_benchmark(benchmarks_list1[[1]], class_types_6th)
benchmarks_list1[[2]] <- process_benchmark(benchmarks_list1[[2]], class_types_6th)
benchmarks_list1[[3]] <- process_benchmark(benchmarks_list1[[3]], class_types_6th)

benchmarks_list1[[4]] <- process_benchmark(benchmarks_list1[[4]], class_types_7th)
benchmarks_list1[[5]] <- process_benchmark(benchmarks_list1[[5]], class_types_7th)
benchmarks_list1[[6]] <- process_benchmark(benchmarks_list1[[6]], class_types_7th)

benchmarks_list1[[7]] <- process_benchmark(benchmarks_list1[[7]], class_types_8th)
benchmarks_list1[[8]] <- process_benchmark(benchmarks_list1[[8]], class_types_8th)
benchmarks_list1[[9]] <- process_benchmark(benchmarks_list1[[9]], class_types_8th)

# View an example from the processed list
print(benchmarks_list1[[1]])
```



