---
title: "Data621_Project"
author: "Saloua Daouki"
date: "2024-12-10"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## # 2023-2024 Science Benchmark Data:

**Research Question:** What are the trends in student performance on science benchmark assessments among 6th, 7th, and 8th graders at a Brooklyn school over the academic year? How can regression modeling help identify and forecast performance patterns, highlighting specific questions or standards that show consistent struggles or growth across three benchmarks, and how accurately can these trends predict future performance?

## 1. Load necessary libraries:

This part sets the foundation for the analysis, making sure all necessary tools are in place before moving on to data cleaning, exploration, and visualization. Without loading these libraries, the subsequent analysis wouldn’t be possible, so it's a crucial preparatory step.

```{r }
# Load necessary libraries for data analysis and visualization
library(readxl)   # For reading Excel files
library(purrr)
library(dplyr)    # For data manipulation and transformation
library(tidyr)    # For tidying data
library(naniar)   # For visualizing and handling missing values
library(visdat)   # For creating heatmaps for missing values
library(ggplot2)  # For data visualization
library(DT)       # For creating interactive data tables in R
library(reshape2) 
library(knitr)
library(corrplot)  # For correlation plots
library(stringr)  # Load stringr for string extraction
library(gt)
library(plotly) #allow the plots to be zoomed into for better inspection without sacrificing plot quality.
```

## 2. Explore the data

### 2.1 Load the data

This section establishes a solid base for my future exploration, cleaning, and analysis of student performance trends. I am now prepared to dive deeper into identifying trends and insights across benchmarks.

```{r}
# Path to the Excel file
file_path1 <- "/Users/salouadaouki/Desktop/Data 621/Data-621-Repo-Assignemnt/Data621_Project/2023 - 2024 SCIENCE BENCHMARK DATA  copy.csv"
```





