---
title: "Untitled"
author: "Carol Campbell", "Souleymane Doumbia", "Kossi Akplaka", "Saloua Daouki"
date: "2023-10-21"
output: html_document
---

# Loading libraries:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction:

The goal of this project is to find and use data to answer the question "Which are the most valued data science. Our group members are; Carol Campbell, Kossi Akplaka, Souleymane Doumbia and Saloua Daouki.

# Loading data:

```{r }
url3 <- 'https://raw.githubusercontent.com/Group-Project-Data-607/Project3/main/Data_Science_job.csv'
skills <- read.csv(file = url3)
head(skills)
```

# Tidying the data:

## Data Jobs:

```{r subsetting the data}
required_skills <- subset(df_skills, select = c('title', 'company_name', 'location', 'Job.posting.location', 'description', 'schedule_type', 'Required.skills'))
colnames(required_skills)[7] ="skills"
head(required_skills)
```

```{r subsetting the data again to pick all data jobs}
skills_subset <- subset(required_skills, grepl("data", required_skills$title, ignore.case = TRUE))
```

```{r extracting hard skills from skills column}
freq_hardskills<- skills_subset %>%
    mutate(R = grepl("r|ggplot", skills, ignore.case=TRUE)) %>%
    mutate(python = grepl("Python|pandas|numpy|matplotlib", skills, ignore.case=TRUE)) %>%
    mutate(SQL = grepl("SQL|postgresql|mysql|nosql", skills, ignore.case=TRUE)) %>%
    mutate(Microsoft_Suite = grepl("excel|word|power point|spreadsheet", skills, ignore.case=TRUE)) %>%
    mutate(Azure = grepl("azur", skills, ignore.case=TRUE)) %>%
    mutate(Power_bi = grepl("power bi|power_bi", skills, ignore.case=TRUE)) %>%
    mutate(Tableau = grepl("tableau", skills, ignore.case=TRUE)) %>%
    mutate(Matlab = grepl("matlab", skills, ignore.case=TRUE)) %>%
    mutate(Java = grepl("java|javascript", skills, ignore.case=TRUE)) %>%
    mutate(gcp = grepl("gcp", skills, ignore.case=TRUE)) %>%
    mutate(cplusplus = grepl("c", skills, ignore.case=TRUE)) %>%
    mutate(Jupyter = grepl("jupyter", skills, ignore.case=TRUE)) %>%
    mutate(Visio = grepl("visio", skills, ignore.case=TRUE)) %>%
    mutate(Git = grepl("git", skills, ignore.case=TRUE)) %>%
    mutate(Machine_learning = grepl("machine learning|hadoop", description, ignore.case=TRUE))%>%
    mutate(Cloud = grepl("aws|snowflake", description, ignore.case=TRUE))%>%
    mutate(Other_skills = grepl("spss|sas|ssis|redshift|crystal|looker|airflow|spark|go|linus|unix|rust|alteryx|microstrategy", skills, ignore.case=TRUE)) %>%
  select(title, R, python, SQL, Microsoft_Suite, Azure, Power_bi, Tableau, Matlab, Java, gcp, cplusplus,Jupyter,Visio, Git, Machine_learning, Cloud, Other_skills)
freq_hardskills
```

```{r extracting soft skills from description column}
freq_softskills<- skills_subset %>%
    mutate(Communication = grepl("communicat", description, ignore.case=TRUE)) %>%
    mutate(Collaborative = grepl("collaborat|group|team", description, ignore.case=TRUE)) %>%
    mutate(TimeManagement = grepl("manage", description, ignore.case=TRUE)) %>%
    mutate(Leadership = grepl("leader", description, ignore.case=TRUE)) %>%
    mutate(Remote = grepl("remote", description, ignore.case=TRUE)) %>%
    mutate(Experienced = grepl("experience", description, ignore.case=TRUE)) %>%
    mutate(Visualization = grepl("visualization", description, ignore.case=TRUE)) %>%
    mutate(Problem_solving = grepl("problem solving", description, ignore.case=TRUE)) %>%      
    mutate(Critical_analysis = grepl("critical", description, ignore.case=TRUE)) %>%
    mutate(Confidence = grepl("confiden", description, ignore.case=TRUE)) %>%
  select(title, Communication, Collaborative, TimeManagement, Leadership, Remote, Experienced, Visualization, Problem_solving, Critical_analysis, Confidence)
freq_softskills
```

## Data Jobs Analysis:

```{r counting the frequency for every hard skill}
hardskills <- freq_hardskills %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
hardskills
```

```{r counting the frequency for every other skill}
softskills <- freq_softskills %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
softskills
```

```{r hard skill plot}
ggplot(hardskills,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="orange") + xlab('') + ylab('Frequency') + labs(title='Data Jobs Hard Skills') + coord_flip()
```

```{r other skill plot}
ggplot(softskills,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="blue") + xlab('') + ylab('Frequency') + labs(title='Data Jobs Required Soft Skills') + coord_flip()
```

## Data Science Jobs

```{r subsetting the data to get data science only}
skills_subset_science <- subset(required_skills, grepl("data science", required_skills$description, ignore.case = TRUE))
```

```{r hard skills for data science jobs}
freq_hardskills_science<- skills_subset_science %>%
    mutate(R = grepl("r|ggplot", skills, ignore.case=TRUE)) %>%
    mutate(python = grepl("Python|pandas|numpy|matplotlib", skills, ignore.case=TRUE)) %>%
    mutate(SQL = grepl("SQL|postgresql|mysql|nosql", skills, ignore.case=TRUE)) %>%
    mutate(Microsoft_Suite = grepl("excel|word|power point|spreadsheet|outlook|sharepoint|vba", skills, ignore.case=TRUE)) %>%
    mutate(Azure = grepl("azur", skills, ignore.case=TRUE)) %>%
    mutate(Power_bi = grepl("power bi|power_bi", skills, ignore.case=TRUE)) %>%
    mutate(Tableau = grepl("tableau", skills, ignore.case=TRUE)) %>%
    mutate(Matlab = grepl("matlab", skills, ignore.case=TRUE)) %>%
    mutate(Java = grepl("java|javascript", skills, ignore.case=TRUE)) %>%
    mutate(gcp = grepl("gcp", skills, ignore.case=TRUE)) %>%
    mutate(cplusplus = grepl("c", skills, ignore.case=TRUE)) %>%
    mutate(Jupyter = grepl("jupyter", skills, ignore.case=TRUE)) %>%
    mutate(Visio = grepl("visio", skills, ignore.case=TRUE)) %>%
    mutate(Git = grepl("git", skills, ignore.case=TRUE)) %>%
    mutate(Machine_learning = grepl("machine learning|hadoop", description, ignore.case=TRUE))%>%
    mutate(Cloud = grepl("aws|snowflake", description, ignore.case=TRUE))%>%
    mutate(Other_skills = grepl("spss|sas|ssis|redshift|crystal|looker|airflow|spark|go|linus|unix|rust|alteryx|microstrategy", skills, ignore.case=TRUE)) %>%
  select(title, R, python, SQL, Microsoft_Suite, Azure, Power_bi, Tableau, Matlab, Java, gcp, cplusplus,Jupyter,Visio, Git, Machine_learning, Cloud, Other_skills)
freq_hardskills_science
```


```{r soft skills for data science jobs}
freq_softskills_science<- skills_subset_science %>%
    mutate(Communication = grepl("communicat", description, ignore.case=TRUE)) %>%
    mutate(Collaborative = grepl("collaborat|group|team", description, ignore.case=TRUE)) %>%
    mutate(TimeManagement = grepl("manage", description, ignore.case=TRUE)) %>%
    mutate(Leadership = grepl("leader", description, ignore.case=TRUE)) %>%
    mutate(Remote = grepl("remote", description, ignore.case=TRUE)) %>%
    mutate(Experienced = grepl("experience", description, ignore.case=TRUE)) %>%
    mutate(Visualization = grepl("visualization", description, ignore.case=TRUE)) %>%
    mutate(Problem_solving = grepl("problem solving", description, ignore.case=TRUE)) %>%      
    mutate(Critical_analysis = grepl("critical", description, ignore.case=TRUE)) %>%
    mutate(Confidence = grepl("confiden", description, ignore.case=TRUE)) %>%
  select(title, Communication, Collaborative, TimeManagement, Leadership, Remote, Experienced, Visualization, Problem_solving, Critical_analysis, Confidence)
freq_softskills_science
```

## Data Science Jobs Analysis:

```{r counting the frequencies of data science jobs }
hardskills_science <- freq_hardskills_science %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
hardskills_science
```

```{r plotting the hard skills for data science}
ggplot(hardskills_science,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="red") + xlab('') + ylab('Frequency') + labs(title='Data Science Required Hard Skills') + coord_flip()
```
```{r}
softskills_science <- freq_softskills_science %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
```

```{r plotting the soft skills for data science}
ggplot(softskills_science,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="lightblue") + xlab('') + ylab('Frequency') + labs(title='Data Science Required Soft Skills') + coord_flip()
```

## Data Engineering Jobs

```{r subsetting the data to get data Engineering only}
skills_subset_engineer <- subset(required_skills, grepl("data engineer", required_skills$description, ignore.case = TRUE))
```

```{r creating a subset of the dataframe with hard skills}
freq_hardskills_engineer<- skills_subset_engineer %>%
    mutate(R = grepl("r|ggplot", skills, ignore.case=TRUE)) %>%
    mutate(python = grepl("Python|pandas|numpy|matplotlib", skills, ignore.case=TRUE)) %>%
    mutate(SQL = grepl("SQL|postgresql|mysql|nosql", skills, ignore.case=TRUE)) %>%
    mutate(Microsoft_Suite = grepl("excel|word|power point|spreadsheet|outlook|sharepoint|vba", skills, ignore.case=TRUE)) %>%
    mutate(Azure = grepl("azur", skills, ignore.case=TRUE)) %>%
    mutate(Power_bi = grepl("power bi|power_bi", skills, ignore.case=TRUE)) %>%
    mutate(Tableau = grepl("tableau", skills, ignore.case=TRUE)) %>%
    mutate(Matlab = grepl("matlab", skills, ignore.case=TRUE)) %>%
    mutate(Java = grepl("java|javascript", skills, ignore.case=TRUE)) %>%
    mutate(gcp = grepl("gcp", skills, ignore.case=TRUE)) %>%
    mutate(cplusplus = grepl("c", skills, ignore.case=TRUE)) %>%
    mutate(Jupyter = grepl("jupyter", skills, ignore.case=TRUE)) %>%
    mutate(Visio = grepl("visio", skills, ignore.case=TRUE)) %>%
    mutate(Git = grepl("git", skills, ignore.case=TRUE)) %>%
    mutate(Machine_learning = grepl("machine learning|hadoop", description, ignore.case=TRUE))%>%
    mutate(Cloud = grepl("aws|snowflake", description, ignore.case=TRUE))%>%
    mutate(Other_skills = grepl("spss|sas|ssis|redshift|crystal|looker|airflow|spark|go|linus|unix|rust|alteryx|microstrategy", skills, ignore.case=TRUE)) %>%
  select(title, R, python, SQL, Microsoft_Suite, Azure, Power_bi, Tableau, Matlab, Java, gcp, cplusplus,Jupyter,Visio, Git, Machine_learning, Cloud, Other_skills)
freq_hardskills_engineer
```


```{r subset of the data to include only soft skills_eng}
freq_softskills_engineer<- skills_subset_engineer %>%
    mutate(Communication = grepl("communicat", description, ignore.case=TRUE)) %>%
    mutate(Collaborative = grepl("collaborat|group|team", description, ignore.case=TRUE)) %>%
    mutate(TimeManagement = grepl("manage", description, ignore.case=TRUE)) %>%
    mutate(Leadership = grepl("leader", description, ignore.case=TRUE)) %>%
    mutate(Remote = grepl("remote", description, ignore.case=TRUE)) %>%
    mutate(Experienced = grepl("experience", description, ignore.case=TRUE)) %>%
    mutate(Visualization = grepl("visualization", description, ignore.case=TRUE)) %>%
    mutate(Problem_solving = grepl("problem solving", description, ignore.case=TRUE)) %>%      
    mutate(Critical_analysis = grepl("critical", description, ignore.case=TRUE)) %>%
    mutate(Confidence = grepl("confiden", description, ignore.case=TRUE)) %>%
  select(title, Communication, Collaborative, TimeManagement, Leadership, Remote, Experienced, Visualization, Problem_solving, Critical_analysis, Confidence)
freq_softskills_engineer
```

## Data Engineering Jobs Analysis:

```{r counting the frequencies of hard skills for data engineer jobs }
hardskills_engineer <- freq_hardskills_engineer %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
hardskills_engineer
```

```{r plotting the hard skills for data engineer}
ggplot(hardskills_engineer,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="magenta") + xlab('') + ylab('Frequency') + labs(title='Data Engineering Required Hard Skills') + coord_flip()
```

```{r counting the frequency of soft skills for data engineer}
softskills_engineer <- freq_softskills_engineer %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
```

```{r plotting the soft skills for data engineer}
ggplot(softskills_engineer,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="lightblue") + xlab('') + ylab('Frequency') + labs(title='Data Engineering Required Soft Skills') + coord_flip()
```

## Data Analyst Jobs

```{r subsetting the data to get data Analyst only}
skills_subset_analyst <- subset(required_skills, grepl("data analyst", required_skills$description, ignore.case = TRUE))
```

```{r creating a subset of the dataframe with hard skills for analyst}
freq_hardskills_analyst<- skills_subset_analyst %>%
    mutate(R = grepl("r|ggplot", skills, ignore.case=TRUE)) %>%
    mutate(python = grepl("Python|pandas|numpy|matplotlib", skills, ignore.case=TRUE)) %>%
    mutate(SQL = grepl("SQL|postgresql|mysql|nosql", skills, ignore.case=TRUE)) %>%
    mutate(Microsoft_Suite = grepl("excel|word|power point|spreadsheet|outlook|sharepoint|vba", skills, ignore.case=TRUE)) %>%
    mutate(Azure = grepl("azur", skills, ignore.case=TRUE)) %>%
    mutate(Power_bi = grepl("power bi|power_bi", skills, ignore.case=TRUE)) %>%
    mutate(Tableau = grepl("tableau", skills, ignore.case=TRUE)) %>%
    mutate(Matlab = grepl("matlab", skills, ignore.case=TRUE)) %>%
    mutate(Java = grepl("java|javascript", skills, ignore.case=TRUE)) %>%
    mutate(gcp = grepl("gcp", skills, ignore.case=TRUE)) %>%
    mutate(cplusplus = grepl("c", skills, ignore.case=TRUE)) %>%
    mutate(Jupyter = grepl("jupyter", skills, ignore.case=TRUE)) %>%
    mutate(Visio = grepl("visio", skills, ignore.case=TRUE)) %>%
    mutate(Git = grepl("git", skills, ignore.case=TRUE)) %>%
    mutate(Machine_learning = grepl("machine learning|hadoop", description, ignore.case=TRUE))%>%
    mutate(Cloud = grepl("aws|snowflake", description, ignore.case=TRUE))%>%
    mutate(Other_skills = grepl("spss|sas|ssis|redshift|crystal|looker|airflow|spark|go|linus|unix|rust|alteryx|microstrategy", skills, ignore.case=TRUE)) %>%
  select(title, R, python, SQL, Microsoft_Suite, Azure, Power_bi, Tableau, Matlab, Java, gcp, cplusplus,Jupyter,Visio, Git, Machine_learning, Cloud, Other_skills)
freq_hardskills_analyst
```


```{r soft skills data for data analyst}
freq_softskills_analyst<- skills_subset_analyst %>%
    mutate(Communication = grepl("communicat", description, ignore.case=TRUE)) %>%
    mutate(Collaborative = grepl("collaborat|group|team", description, ignore.case=TRUE)) %>%
    mutate(TimeManagement = grepl("manage", description, ignore.case=TRUE)) %>%
    mutate(Leadership = grepl("leader", description, ignore.case=TRUE)) %>%
    mutate(Remote = grepl("remote", description, ignore.case=TRUE)) %>%
    mutate(Experienced = grepl("experience", description, ignore.case=TRUE)) %>%
    mutate(Visualization = grepl("visualization", description, ignore.case=TRUE)) %>%
    mutate(Problem_solving = grepl("problem solving", description, ignore.case=TRUE)) %>%      
    mutate(Critical_analysis = grepl("critical", description, ignore.case=TRUE)) %>%
    mutate(Confidence = grepl("confiden", description, ignore.case=TRUE)) %>%
  select(title, Communication, Collaborative, TimeManagement, Leadership, Remote, Experienced, Visualization, Problem_solving, Critical_analysis, Confidence)
freq_softskills_analyst
```

## Data Analyst Jobs Analysis:

```{r counting the frequencies of hard skills for data analyst jobs }
hardskills_analyst <- freq_hardskills_analyst %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
hardskills_analyst
```

```{r plotting the hard skills for data analyst}
ggplot(hardskills_analyst,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="yellow") + xlab('') + ylab('Frequency') + labs(title='Data Analyst Required Hard Skills') + coord_flip()
```

```{r counting the frequency of soft skills for data analyst}
softskills_analyst <- freq_softskills_analyst %>%
  select(-1)%>%
  summarise_all(sum)%>%
  gather(skill, freq)%>%
  arrange(desc(freq))
```

```{r plotting the soft skills for data analyst}
ggplot(softskills_analyst,aes(x=reorder(skill, freq), y=freq)) + geom_bar(stat='identity',fill="brown") + xlab('') + ylab('Frequency') + labs(title='Data Analyst Required Soft Skills') + coord_flip()
```



# Conclusion:
