---
title: "DATA 621 Assignment 2"
author: "Warner Alexis"
date: "2024-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Overview

In this homework assignment, you will work through various classification metrics. You will be asked to create functions in R to carry out the various calculations. You will also investigate some functions in packages that will let you obtain the equivalent results. Finally, you will create graphical output that also can be used to evaluate the output of classification models, such as binary logistic regression.


The data set has three key columns we will use:
> class: the actual class for the observation

> scored.class: the predicted class for the observation (based on a threshold of 0.5)

> scored.probability: the predicted probability of success for the observation

Use the table() function to get the raw confusion matrix for this scored dataset. Make sure you understand the output. In particular, do the rows represent the actual or predicted class? The columns?

```{r}
# Load library
library(DataExplorer)
library(dplyr)
library(caret)
library(MASS)

# Laod the dataset 
datac <- read.csv('classification-output-data.csv', stringsAsFactors = F)
head(datac)
str(datac)
```




Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the accuracy of the predictions.

> $ğ´ğ‘ğ‘ğ‘¢ğ‘Ÿğ‘ğ‘ğ‘¦= ğ‘‡ğ‘ƒ+ğ‘‡ğ‘ğ‘‡ğ‘ƒ+ğ¹ğ‘ƒ+ğ‘‡ğ‘+ğ¹ğ‘$ 

4. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the classification error rate of the predictions.

 $ğ¶ğ‘™ğ‘ğ‘ ğ‘ ğ‘–ğ‘“ğ‘–ğ‘ğ‘ğ‘¡ğ‘–ğ‘œğ‘› ğ¸ğ‘Ÿğ‘Ÿğ‘œğ‘Ÿ ğ‘…ğ‘ğ‘¡ğ‘’= (ğ¹ğ‘ƒ+ğ¹ğ‘)/(ğ‘‡ğ‘ƒ+ğ¹ğ‘ƒ+ğ‘‡ğ‘+ğ¹ğ‘) $
 
 
 
```{r}
# create a model 
conf_matrix <- table(datac$class, datac$scored.class)
conf_matrix
conf_matrix[1,1]


## Let do it on caret 
data <- datac
data$class <- as.factor(data$class)
data$scored.class <- as.factor(data$scored.class)
# use function in caret package to do it 
confusion <- confusionMatrix(data$scored.class, data$class, positive= '1')

get_Accuracy <- function(confusion_matrix){
  TN <- confusion_matrix[1,1] # True Negative
  FP <- confusion_matrix[1,2] # False Positive
  FN <- confusion_matrix[2,1] # False Negative
  TP <- confusion_matrix[2,2] # True Positive
  
  # calculate accuracy 
  accuracy <- (TP + TN) / (TP + TN + FP + FN)
  
  # Classification Error Rate 
  class_err_rate <- (FP + FN) / (TP + FP + TN + FN)
  
  # Use cat() to print multiple values
  cat('The accuracy is: ', accuracy, '\n')
  cat('The Classification Error Rate is: ', class_err_rate, '\n')
  cat('The sum of both accuracy and classification error is ', sum(accuracy, class_err_rate))
}

get_Accuracy(conf_matrix)
precision(data = as.factor(datac$scored.class), 
          reference = as.factor(datac$class),
          positive = '0')

```
 
 
5 Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the precision of the predictions.

$ğ‘ƒğ‘Ÿğ‘’ğ‘ğ‘–ğ‘ ğ‘–ğ‘œğ‘›= ğ‘‡ğ‘ƒ/(ğ‘‡ğ‘ƒ+ğ¹ğ‘ƒ)$




```{r}

get_precision <- function(confusion_matrix){
  TN <- confusion_matrix[1,1] # True Negative
  FP <- confusion_matrix[1,2] # False Positive
  FN <- confusion_matrix[2,1] # False Negative
  TP <- confusion_matrix[2,2] # True Positive
  
  # calculate accuracy 
  precision <-  (TP)/ (TP + FP)
  
  
  
  # Use cat() to print multiple values
  cat('The precision is: ', precision, '\n')
  
}
get_precision(conf_matrix)
# other way to calculated 
caret_precision <- confusion$byClass['Precision']
caret_precision
```


5 Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the sensitivity of the predictions. Sensitivity is also known as recall.

$ğ‘†ğ‘’ğ‘›ğ‘ ğ‘–ğ‘¡ğ‘–ğ‘£ğ‘–ğ‘¡ğ‘¦= ğ‘‡ğ‘ƒ/(ğ‘‡ğ‘ƒ+ğ¹ğ‘)$


```{r}
# sensitivity 
get_sensitivity <- function(confusion_matrix){
  TN <- confusion_matrix[1,1] # True Negative
  FP <- confusion_matrix[1,2] # False Positive
  FN <- confusion_matrix[2,1] # False Negative
  TP <- confusion_matrix[2,2] # True Positive
  
  # calculate accuracy 
  sensitivity <-  (TP)/ (TP + FN)
  
  
  
  # Use cat() to print multiple values
  cat('The Sensitivity is: ', sensitivity, '\n')
  
}
get_sensitivity(conf_matrix)
# other way to calculated 
caret_sensitivity <- confusion$byClass['Sensitivity']
caret_sensitivity
```

6 Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the specificity of the predictions.
$ ğ‘†ğ‘ğ‘’ğ‘ğ‘–ğ‘“ğ‘–ğ‘ğ‘–ğ‘¡ğ‘¦= ğ‘‡ğ‘/(Tğ‘+ğ¹ğ‘ƒ)$

```{r}

# specificity
get_specificity <- function(confusion_matrix){
  TN <- confusion_matrix[1,1] # True Negative
  FP <- confusion_matrix[1,2] # False Positive
  FN <- confusion_matrix[2,1] # False Negative
  TP <- confusion_matrix[2,2] # True Positive
  
  # calculate specificity
  specificity <-  (TN)/ (TN + FP)
  
  
  
  # Use cat() to print multiple values
  cat('The specificity is: ', specificity, '\n')
  
}
get_specificity(conf_matrix)
# other way to calculated 
caret_specificity <- confusion$byClass['Specificity']
caret_specificity
```




8 Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the F1 score of the predictions.

$ğ¹1 ğ‘†ğ‘ğ‘œğ‘Ÿğ‘’= 2(ğ‘ƒğ‘Ÿğ‘’ğ‘ğ‘–ğ‘ ğ‘–ğ‘œğ‘›Ã—ğ‘†ğ‘’ğ‘›ğ‘ ğ‘–ğ‘¡ğ‘–ğ‘£ğ‘–ğ‘¡ğ‘¦)/(ğ‘ƒğ‘Ÿğ‘’ğ‘ğ‘–ğ‘ ğ‘–ğ‘œğ‘›+ğ‘†ğ‘’ğ‘›ğ‘ ğ‘–ğ‘¡ğ‘–ğ‘£ğ‘–ğ‘¡ğ‘¦)$

```{r}

# F1 score

get_f1_score <- function(confusion_matrix){
  TN <- confusion_matrix[1,1] # True Negative
  FP <- confusion_matrix[1,2] # False Positive
  FN <- confusion_matrix[2,1] # False Negative
  TP <- confusion_matrix[2,2] # True Positive
  
  # calculate specificity
  # calculate precision 
  precision <-  (TP)/ (TP + FP)
  # calculate sensitivity 
  sensitivity <-  (TP)/ (TP + FN)
  
  # calculate f1_score
  f1_score <- 2*(precision * sensitivity)/(precision + sensitivity)
  
  # Use cat() to print multiple values
  cat('The F1 score is: ', f1_score, '\n')
  
}
get_f1_score(conf_matrix)
```

